---
title: "On benchmarking gene set enrichment analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE)
knitr::opts_chunk$set(cache=FALSE)
```

## Introduction

Systematic evaluation of gene set enrichment analysis methods 
is a much needed but challenging task. 
Recently Geistlinger and colleagues presented a framework
[GSEABenchmarkeR](https://bioconductor.org/packages/GSEABenchmarkeR)
for extensible and reproducible gene set enrichment benchmarking. 
Here we show that the MalaCards-based phenotype relevance score,
one of the key framework's component,
turns out to be a poor metric for comparing quality of enrichment results.
We demonstrate this by showing that a baseline method that simply orders genes sets by their 
size significantly outperforms all other considered methods, including frequently used
in practice ORA and GSEA methods.


### Loading libraries

```{r message=FALSE}
library(GSEABenchmarkeR)
library(EnrichmentBrowser)
library(data.table)
library(reshape2)
library(ggplot2)
library(ggpubr)

source("methods/utils.R")
```

### Microarray and TCGA ids

```{r message=FALSE}
geo2kegg <- loadEData("geo2kegg")
ma.ids <- names(geo2kegg)

rseq.ids <- c("BLCA", "BRCA", "COAD", "HNSC", "KICH", 
              "KIRC", "KIRP", "LIHC", "LUAD", "LUSC", 
              "PRAD", "READ", "STAD", "THCA", "UCEC")
```

### Reading rankings from the article

```{r}
ea.methods <- sbeaMethods()[1:10]

rseq.kegg.ranks <- readResults("inpdata/TCGA/GSE62944_matched_vst/kegg/", rseq.ids,
                               methods = ea.methods, type="ranking")
rseq.go.ranks <- readResults("inpdata/TCGA/GSE62944_matched_vst/gobp/", rseq.ids,
                               methods = ea.methods, type="ranking")

ma.kegg.ranks <- readResults("inpdata/GEO2KEGG/kegg/perm1k/", ma.ids, 
                             methods = ea.methods, type="ranking")
ma.go.ranks <- readResults("inpdata/GEO2KEGG/go_bp/", ma.ids, 
                             methods = ea.methods, type="ranking")
```

## Benchmarking

Here we introduce a baseline enrichment method `GSS`, which ranks gene sets based only on their size.

### Adding geneset-size based ranking

```{r}
gss.rseq.kegg <- lapply(rseq.ids, function(rseqId){
    oraRanking <- as.data.table(rseq.kegg.ranks$ora[[rseqId]])
    gssRanking <- oraRanking[, .(GENE.SET, NR.GENES)]
    gssRanking[, PVAL := 1 / NR.GENES]
    return(as.data.frame(gssRanking))
})
names(gss.rseq.kegg) <- rseq.ids

gss.rseq.go <- lapply(rseq.ids, function(rseqId){
    oraRanking <- as.data.table(rseq.go.ranks$ora[[rseqId]])
    gssRanking <- oraRanking[, .(GENE.SET, NR.GENES)]
    gssRanking[, PVAL := 1 / NR.GENES]
    return(as.data.frame(gssRanking))
})
names(gss.rseq.go) <- rseq.ids

gss.ma.kegg <- lapply(ma.ids, function(maId){
    oraRanking <- as.data.table(ma.kegg.ranks$ora[[maId]])
    gssRanking <- oraRanking[, .(GENE.SET, NR.GENES)]
    gssRanking[, PVAL := 1 / NR.GENES]
    return(as.data.frame(gssRanking))
})
names(gss.ma.kegg) <- ma.ids

gss.ma.go <- lapply(ma.ids, function(maId){
    oraRanking <- as.data.table(ma.go.ranks$ora[[maId]])
    gssRanking <- oraRanking[, .(GENE.SET, NR.GENES)]
    gssRanking[, PVAL := 1 / NR.GENES]
    return(as.data.frame(gssRanking))
})
names(gss.ma.go) <- ma.ids

ma.kegg.ranks$gss <- gss.ma.kegg
ma.go.ranks$gss <- gss.ma.go
rseq.kegg.ranks$gss <- gss.rseq.kegg
rseq.go.ranks$gss <- gss.rseq.go
```

### Reading mala cards

```{r}
data.dir <- system.file("extdata", package="GSEABenchmarkeR")
mala.kegg.file <- file.path(data.dir, "malacards", "KEGG.rds")
mala.go.file <- file.path(data.dir, "malacards", "GO_BP.rds")
mala.kegg <- readRDS(mala.kegg.file)
mala.go <- readRDS(mala.go.file)

d2d.file <- file.path(data.dir, "malacards", "GseId2Disease.txt")
d2d.map <- readDataId2diseaseCodeMap(d2d.file)

d2d.tcga <- rseq.ids
names(d2d.tcga) <- rseq.ids
```

### Evaluating relative scores

```{r}
ma.kegg.rel.sets <- evalRelevance(ma.kegg.ranks, mala.kegg, d2d.map)
ma.go.rel.sets <- evalRelevance(ma.go.ranks, mala.go, d2d.map)

rseq.kegg.rel.sets <- evalRelevance(rseq.kegg.ranks, mala.kegg, d2d.tcga)
rseq.go.rel.sets <- evalRelevance(rseq.go.ranks, mala.go, d2d.tcga)
```

### Plot results

```{r}
facetplot(ma.kegg.rel.sets, ma.go.rel.sets, 
          rseq.kegg.rel.sets, rseq.go.rel.sets, 
          ylab="% optimal relevance score", vline = NA)
```

## Overall difference between `GSS` and other methods

```{r}
vapply(setdiff(ea.methods, "gss"), 
        function(m) testMethods("gss", m), 
        numeric(5))
```


## Session info

```{r}
sessionInfo()
```
